---
title: "Linear Regression from Scratch"
format: 
  quarto: true
execute:
  echo: true
  warning: false
  error: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Libraries to use: pandas, numpy, plotly
```

# 📈 Linear Regression: Modeling Linear Relationships

Linear Regression is one of the most fundamental supervised learning algorithms. Its goal is to model the relationship between a dependent variable ($y$) and one or more independent variables ($x$) by fitting a straight line (or a hyperplane in multiple dimensions) to the observed data. The model assumes a linear relationship, represented by the equation: $y = \theta_0 + \theta_1 x$.

## 🏛️ General Architecture

The architecture is straightforward: we use a linear hypothesis function and minimize the cost of the predictions using an optimization technique, which is typically Gradient Descent.

*Diagram Concept: A scatter plot of data points with a single, best-fit line drawn through them.*

**Process Flow:**
1.  Initialize Parameters ($\theta_0$ and $\theta_1$).
2.  Define the Linear Hypothesis: $\hat{y} = \theta_0 + \theta_1 x$.
3.  Calculate Cost (MSE) based on the difference between $\hat{y}$ and the actual $y$.
4.  Use Gradient Descent to iteratively adjust $\theta_0$ and $\theta_1$ to minimize the cost.

## 🛠️ Step-by-Step Implementation (Python/NumPy)

This implementation will focus on building the linear model and training it using the Gradient Descent technique already established.

```{r setup, include=FALSE}
# Ensure necessary libraries are installed/available in the environment
library(pandas)
library(numpy)
library(plotly)
```

### 1. Data Preparation
We will use a synthetic dataset to demonstrate the model fitting process.

```{python data-prep}
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

# Generate synthetic data (y = 3x + 5 + noise)
np.random.seed(10)
X = 3 * np.random.rand(100, 1)
y = 5 + 3 * X + np.random.randn(100, 1)

# Convert to DataFrame for easier handling
data = pd.DataFrame({'X': X.flatten(), 'y': y.flatten()})
print(data.head())
```

### 2. Model Definition & Cost Function
The model is $\hat{y} = \theta_0 + \theta_1 x$. The cost function remains MSE.

```{python cost-function}
import numpy as np

def compute_cost(X, y, theta):
    """
    Compute Mean Squared Error (MSE) cost function.
    J(θ) = (1/(2m)) * Σ(h_θ(x) - y)²
    """
    m = len(y)
    predictions = X.dot(theta)
    cost = (1 / (2 * m)) * np.sum((predictions - y)**2)
    return cost
```

### 3. Training Loop (Gradient Descent Application)
We reuse the core GD logic to train the linear model.

```{python training-loop}
import numpy as np

def compute_gradient(X, y, theta):
    """
    Compute gradient of the cost function.
    ∂J(θ) = (1/m) * Σ(h_θ(x) - y) * x_j
    """
    m = len(y)
    predictions = X.dot(theta)
    errors = predictions - y
    gradient = (1 / m) * X.T.dot(errors)
    return gradient


def gradient_descent(X, y, theta_init, learning_rate, n_iterations):
    """
    Run gradient descent algorithm.
    θ = θ - α * ∇J(θ)
    """
    theta = theta_init.copy()
    cost_history = []
    
    for iteration in range(n_iterations):
        # Compute cost and gradient
        cost = compute_cost(X, y, theta)
        gradient = compute_gradient(X, y, theta)
        
        # Store cost history
        cost_history.append(cost)
        
        # Update parameters
        theta = theta - learning_rate * gradient
        
    return theta, cost_history
```

### 4. Visualization of Fit
We plot the final learned line against the data.

```{python visualization-fit}
import matplotlib.pyplot as plt

# Function to plot the data and fitted line
# Assuming final_theta is available from the training loop

def plot_fit(X, y, theta):
    """Plot the original data points and the fitted line."""
    plt.figure(figsize=(10, 5))
    
    # Plot original data
    plt.scatter(X.flatten(), y.flatten(), c='blue', alpha=0.6, label='Data Points')
    
    # Plot fitted line
    x_line = np.linspace(0, 3, 100).reshape(-1, 1)
    y_line = theta.dot(x_line)
    plt.plot(x_line, y_line, r='r', linewidth=2, label=f'Fitted Line: y = {theta[0,0]:.3f} + {theta[1,0]:.3f}*x')
    
    plt.title('Data and Fitted Linear Model', fontsize=14, fontweight='bold')
    plt.xlabel('x', fontsize=12)
    plt.ylabel('y', fontsize=12)
    plt.legend()
    plt.grid(True, linestyle='--', alpha=0.7)
    plt.show()
```

## 📝 Summary

Linear Regression provides a simple yet powerful framework for prediction. By deriving the model parameters ($\theta_0, \theta_1$) through optimization, we create a line that best describes the linear trend in the data, allowing us to make informed predictions.
