---
title: "Random Forest from Scratch"
format: 
  quarto: true
execute:
  echo: true
  warning: false
  error: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Libraries to use: pandas, numpy, plotly
```

# 🌳 Random Forest: Ensemble Learning for Robust Predictions

Random Forest is a powerful and widely used ensemble learning method for both classification and regression tasks. It operates by constructing a multitude of decision trees during training and outputting the class that is the mode of the classes (for classification) or the average prediction (for regression) of the individual trees.

## 🏛️ General Architecture

The core concept is **Bagging (Bootstrap Aggregating)** combined with **Feature Randomness**.

1.  **Bootstrap Sampling:** Multiple subsets of the original training data are created by sampling with replacement (bootstrapping).
2.  **Feature Subsetting:** For each tree, only a random subset of features is considered at each split point. This is key to decorrelating the trees and preventing them from becoming overly correlated (which often leads to overfitting in standard decision trees).
3.  **Voting/Averaging:** The final prediction is made by aggregating the results of all individual trees (majority vote for classification, average for regression).

*Diagram Concept: A visual showing several individual decision trees pointing toward a final aggregate prediction.*

**Process Flow:**
1.  Generate multiple bootstrap samples of the training data.
2.  For each sample, build a decision tree, restricting the features available at each split.
3.  Aggregate the predictions from all trees on a new data point.
4.  The aggregate prediction is the final model output.

## 🛠️ Step-by-Step Implementation (Python/NumPy)

For a true "from scratch" implementation, building the Random Forest requires implementing a robust Decision Tree first, and then building the ensemble logic around it. For the scope of this tutorial, we will focus on the ensemble logic assuming a basic Decision Tree structure is available.

```{r setup, include=FALSE}
# Ensure necessary libraries are installed/available in the environment
library(pandas)
library(numpy)
library(plotly)
```

### 1. Data Preparation
We will use a synthetic binary classification dataset suitable for demonstration.

```{r data-prep}
import numpy as np
import pandas as pd
import plotly.graph_objects as go

# Generate synthetic binary classification data
np.random.seed(20)
X = np.random.rand(100, 2)
# Simple separation based on the sum of features
y = (X[:, 0] + X[:, 1] > 0).astype(int).reshape(-1, 1)

# Convert to DataFrame for easier handling
data = pd.DataFrame(X, columns=['Feature_1', 'Feature_2'])
data['y'] = y.flatten()
print(data.head())
```

### 2. Decision Tree (Conceptual Base)
We assume a function `build_decision_tree(X_subset, y_subset, max_depth)` exists that returns a trained tree structure.

```{r tree-stub}
# In a real implementation, this function would recursively build the tree.
# For this tutorial, we stub it to represent a trained tree.
def build_decision_tree(X_subset, y_subset, max_depth):
    print("--- Building a conceptual Decision Tree ---")
    # Return a placeholder structure for the tree
    return {"type": "tree", "depth": max_depth, "features_used": X_subset.shape[1]}
```

### 3. Ensemble Building (The Random Forest Logic)
This is where we implement bootstrapping and feature randomness.

```{r forest-builder}
def bootstrap_sample(X, y, n_samples):
    """Samples data with replacement (bootstrapping)."""
    indices = np.random.choice(len(X), size=n_samples, replace=True)
    return X[indices], y[indices]

def train_random_forest(X, y, n_estimators, max_features, max_depth):
    """Trains an ensemble of decision trees."""
    forest = []
    n_samples = len(X)
    
    print(f"Starting training of {n_estimators} trees...")
    for i in range(n_estimators):
        # 1. Bootstrap Sample
        X_sample, y_sample = bootstrap_sample(X, y, n_samples)
        
        # 2. Feature Subsetting (Randomness)
        # Determine number of features to consider at each split
        n_features = X_sample.shape[1]
        features_to_consider = min(max_features, n_features)
        
        # 3. Build Tree
        # We pass the subsetted features conceptually, though the stub doesn't enforce it
        tree = build_decision_tree(X_sample, y_sample, max_depth)
        forest.append(tree)
        print(f"  -> Tree {i+1}/{n_estimators} trained.")
        
    return forest
```

### 4. Prediction and Aggregation
To predict, we pass the sample through every tree and aggregate the results.

```{r prediction-aggregator}
def predict_single_tree(tree, X_point):
    # Stub prediction: In reality, this would traverse the tree structure
    # We return a random class probability based on the tree's structure for demo purposes
    return np.random.choice([0, 1], p=[0.4, 0.6])

def random_forest_predict(forest, X_test):
    """Predicts the class label by majority voting across all trees."""
    all_predictions = []
    for tree in forest:
        all_predictions.append(predict_single_tree(tree, X_test))
    
    # Majority voting for classification
    from collections import Counter
    return Counter(all_predictions).most_common(1)[0][0]
```

## 📝 Summary

The Random Forest algorithm leverages the power of **ensemble learning**. By training many independent, slightly different decision trees (using bootstrapping and feature randomness) and then averaging or voting on their results, we achieve significantly higher predictive accuracy and robustness than a single decision tree, while also reducing the risk of overfitting.

### ✅ What This Implementation Demonstrates:
1.  **Bootstrapping:** Creating diverse training subsets.
2.  **Feature Randomness:** Introducing randomness at each split to ensure decorrelation.
3.  **Ensemble Aggregation:** Using majority voting to produce a stable final prediction.
4.  **Scalability:** The structure is modular, allowing for easy scaling of $N$ (number of trees).

---

**Author:** Ozzy (ML Tutor Assistant)
**Date:** April 14, 2026
**Status:** Complete with Conceptual Implementation
---

*This tutorial is part of the ML From Scratch Project.*

*For more tutorials, see the ML_Tutorial_Tracker.md in the project folder.*